---
title: "25 SP STT 180 Homework 2"
author: "<Araynah Dover>"
date: "Feb 25 - Mar 14, 2025"
output: 
  html_document:
    toc: true
    number_sections: false
    toc_float: true
    toc_depth: 4
    df_print: paged
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE,
                      comment = NA)
```

**Setting up:**

Load the `tidyverse` package, which includes `dplyr`, `ggplot2`,
`tidyr`, as well as other packages. Load also the `knitr` package.

```{r echo=TRUE}
library(tidyverse)
library(knitr)
```

Homework 2 has two sections. In Section 1 you will use data
visualization and create functions to analyze a dataset. For Section 2
you will read an article, explore the data, validate the claims and come
to your own conclusions.

**General Instructions:**

-   This is an individual assignment. You may consult with others as you
    work on the assignment, but each student should write up a separate
    set of solutions.

-   Rather than creating a new Rmd file, just add your solutions to the
    supplied Rmd file. **Submit both the Rmd file and the resulting HTML
    file to D2L**. Please note if you are compiling to a pdf, change the
    option in the YAML and also delete the .css lines in the file.

-   Except for questions, or parts of questions, that ask for your
    commentary, use R in a code chunk to answer the questions.

-   The code chunk option `echo = TRUE` is specified in the setup code
    chunk at the beginning of the document. Please do not override this
    in your code chunks.

-   A solution will lose points if the Rmd file does not compile. If one
    of your code chunks is causing your Rmd file to not compile, you can
    use the `eval = FALSE` option. Another possibility is to use the
    `error = TRUE` option in the code chunk.

-   All code chunks should be properly names and well formatted to
    receive full credit.

-   Any use of external souces must be properly cited, including AI
    sources. If using AI, a proper citation should include the tool
    used, the prompt chain, and the date accessed. Not including proper
    citations is plagiarism and could result in a 0.

-   The supplied reference HTML document provides some of the expected
    figures and results, which can be consulted to verify your code is
    correct.

-   This Homework is due on **Friday, March 14th, 2025 on or before 11
    pm.**

# Section 1

This section will use the `BirthDataWithRegionColors.csv` data set. This
contains data about births in the United States, collected by the
Centers for Disease Control (CDC). The variables in the data-set a is
subset of the variables in the full data set, and the observations
represent a "small" sample (chosen at random) of slightly over one
million records from the full data set.

## Introduction

Read in the data and convert the data frame to a tibble.

```{r Read Bday Data, echo=TRUE}
birth_data <- read.csv("BirthDataWithRegionColors.csv", header = TRUE)
birth_data <- as_tibble(birth_data)
```

A glimpse of the data:

```{r View Data Attributes}
glimpse(birth_data)
```

The variables in the data set are:

| Variable        | Description                                                                                                              |
|-------------------------------|-----------------------------------------|
| `year`          | the year of the birth                                                                                                    |
| `month`         | the month of the birth                                                                                                   |
| `state`         | the state where the birth occurred, including "DC" for Washington D.C.                                                   |
| `is_male`       | which is `TRUE` if the child is male, `FALSE` otherwise                                                                  |
| `weight_pounds` | the child's birth weight in pounds                                                                                       |
| `mother_age`    | the age of the mother                                                                                                    |
| `child_race`    | race of the child.                                                                                                       |
| `plurality`     | the number of children born as a result of the pregnancy, with 1 representing a single birth, 2 representing twins, etc. |

Combine `dplyr`with `ggplot2` to create graphical displays of the data.
Use `filter`, `group_by`, and `summarize` to build the required data
frame.

### Question 1

```{=html}
<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
```
::: blue
Create a plot of median age of mother versus year, which includes
separate lines for each of the four regions of the country. (Don't
include data for which the region is missing.) The graphic should look
like the following.
:::

```{r median age of mother vs. year(regions)}
median_age_data <- birth_data %>%
  filter(!is.na(region)) %>%  
  group_by(year, region) %>%  
  summarise(median_age = median(mother_age, na.rm = TRUE)) 

ggplot(data = median_age_data, aes(x = year, y = median_age, color = region)) +
  geom_line()
```

### Question 2.1

```{=html}
<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
```
::: blue
Create a graphic of median age of mother versus year, which includes
separate lines for each of the three values of `state_color`. (Don't
include data for which `state_color` is missing.) The graphic should
look like the following. Notice that the colors are different from the
default colors.
:::

```{r median age of mother vs. year(state_color)}
median_age_data <- birth_data %>%
  filter(!is.na(state_color)) %>%  
  group_by(year, state_color) %>%  
  summarise(median_age = median(mother_age, na.rm = TRUE)) 

ggplot(data = median_age_data, aes(x = year, y = median_age, color = state_color)) +
  geom_line() +
  scale_color_manual(values = median_age_data$state_color)
```

### Question 2.2

Write 2-3 sentences comparing and interpreting the figures from Question
1 and Question 2.1.

*Both figures show trends in the median age of mothers over time. The state_color graph reveals differences based on political standing, the region graph shows the geographical differences *

### Question 3

```{=html}
<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
```
::: blue
Create a graphic of median weight of the child versus year, which
includes separate lines for the first two race categories in the data,
white and black, coded as 1 and 2, respectively. The graphic should look
like the following. Notice that the legend is different from the default
legend. You'll want to investigate `scale_color_discrete` to change the
legend.
:::

```{r median weight of the child vs year}
median_weight_data <- birth_data %>%
  filter(!is.na(child_race) & child_race %in% c(1, 2)) %>%  
  group_by(year, child_race) %>%  
  summarise(median_weight = median(weight_pounds, na.rm = TRUE)) 

median_weight_data$child_race <- factor(median_weight_data$child_race, levels = c(1, 2), labels = c("White", "Black"))

ggplot(data = median_weight_data, aes(x = year, y = median_weight, color = child_race)) +
  geom_line() +
  scale_color_discrete(name = "Race of child")
```

### Question 4

```{=html}
<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
```
::: blue
Create a graphic showing side-by-side boxplots of the weight of the
children for the four regions. (Don't include data for which `region` is
missing.) The graphic should look like the following.
:::

```{r Weight of children boxplot}
region_clean <- birth_data %>%
  filter(!is.na(region)) %>%
  group_by(weight_pounds,region)

ggplot(region_clean, aes(region,weight_pounds)) +
  geom_boxplot()
```

### Question 5

```{=html}
<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
```
::: blue
Write a function called `quick_summary` which takes two inputs:\
`x`: A numeric vector\
`group`: A factor vector of the same length as x

and produces a **list** as output which contains the following
elements:\
\
`missing`: The number of missing values in x\
`means`: The means of x for each level of groups.\
`sds`: The standard deviations of x for each level of groups\

Hint: When computing the means and standard deviations, you need to
exclude missing values using `na.rm = TRUE`.
:::

<p>

```{r Summary Function}
quick_summary <- function(x,group){
  
  list_summary <- list()
  
  missing <- sum(is.na(x))
  
  list_summary$missing <- missing
  
  list_summary$means <- tapply(x, group, mean, na.rm = TRUE)
  list_summary$sds <- tapply(x, group, sd, na.rm = TRUE)
  
  return(list_summary)
}
```

Here is an example of the function in action:

```{r Function call}
quick_summary(birth_data$weight_pounds, birth_data$is_male)
```

### Question 6

```{=html}
<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
```
::: blue
Write a function called `binary_summary` which takes two inputs:\
\
`x`: A vector containing the values 0 and 1 (possibly NA)\
`group`: A factor vector of the same length as x

and produces a **list** as output which contains the following
elements:\
\
`missing`: The number of missing values in x\
`prop`: The proportion of 1s in x for each level of groups\
:::

<p>

```{r Binary Summary}
binary_summary <- function(x, group){
  
  summary_list <- list()
  
  summary_list$missing <- sum(is.na(x))
  
  summary_list$prop <- tapply(x, group, mean, na.rm = TRUE) 
  
  return(summary_list)
}
```

Here is an example of the function in action using `multiple_births`,
defined as a binary variable (0 for no and 1 for yes), and `is_male`:

```{r Binary function call}
#First define a new binary variable, multiple_births, using the plurality variable
birth_data$multiple_births <- ifelse(as.factor(birth_data$plurality)==1, 0, 1)

binary_summary(birth_data$multiple_births, birth_data$is_male)
```

# Section 2

Flint is the second poorest city of its size in the United States and
has spent six of the past 15 years in a state of financial emergency.
One of the cost-cutting measures taken by emergency managers was to stop
buying water, sourced from Lake Huron, from the Detroit Water and
Sewerage Department. Instead, Flint would use the Flint River for its
water supply while waiting for a new pipeline to Lake Huron to be
opened. The move was expected to save roughly \$5 million over a period
of two years.

The Flint River supply was switched on in April 2014. Not long after,
problems arose. Flint resident and mother of four LeeAnne Walters
noticed that the water coming out of her taps was orange. More
worryingly, her family's hair was falling out, her preschool sons had
broken out in rashes and one of them had stopped growing.

The orange colour was from iron, but the family's symptoms pointed to a
far more dangerous contaminant: lead. (Langkjaer - Bain 2017)

## Introduction

The data set consists of 271 homes sampled with three water lead
contaminant values at designated time points. The lead content is in
parts per billion (ppb). Additionally, some location data is given about
each home.

To get started, read in the `flint.csv` file using the function
`read.csv`, as was done in the ICA with the cereal data. However, you do
not need to use the `attach` function. The data set has five variables:

-   **id**: sample id number
-   **zip**: zip code in Flint as to the water sample's location
-   **ward**: ward in Flint as to the water sample's location
-   **draw**: water sample at one of three time points
-   **lead**: lead content in parts per billion

Before you get started, read *The murky tale of Flint's deceptive water
data* by Langkjaer - Bain (2017).

```{r read data}
flint <- read.csv("flint.csv")
flint
```

### Question 1

```{=html}
<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
```
::: blue
Select one passage that you found particularly striking (perhaps you
strongly agreed or disagreed with it, perhaps it made you question an
assumption or seemed unclear) in the article and write a 4-5 sentence
paragraph commenting on it.
:::

*The whole article was pretty interesting, one thing that struck me was the misleading data they brought forth to try and make it seem like the drinking water could have not been to blame for the occuring sickness through out the city. As stated in the article flint is one of the poorest cities in the United states in terms if size, I think the elected officals thought they could just get away with the bad drinking water just based off the fact that the city really doesn’t hsve the funds to provide to do their own research in the water*

### Question 2

```{=html}
<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
```
::: blue Using code, how many unique zip codes are in the data set? How
many unique wards are in the data set?

```{r Unique Sum of zip & ward}
sum(unique(flint$zip))
sum(unique(flint$ward))
```

### Question 3

```{=html}
<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
```
::: blue
Ignoring ward 0, which ward appears to have the worst water quality?
Your answer should take into consideration the mean, median, maximum and
standard deviation of lead levels of each ward. Your choice of 'worst
ward' should include justification for why some of these statistics are
more important to consider than others.
:::

```{r Ward water Quality}
flint %>%
  filter(ward != 0) %>% 
  group_by(ward) %>%
  summarise(mean_lead = mean(lead, na.rm = TRUE),
    median_lead = median(lead, na.rm = TRUE),
    max_lead = max(lead, na.rm = TRUE),
    sd_lead = sd(lead, na.rm = TRUE)) %>%
  arrange(desc(mean_lead))
  
```

*Based on the summary of the variables it looks like ward 2 contained the greatest lead levels with an average of 24 and a max of 1051 which is actually devastating*

### Question 4

```{=html}
<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
```
::: blue
Langkjear-Bain (2017) writes at length about the practice of 'drawing'
water before sampling it for lead levels. Compute the median and mean
lead values for each draw. How do they compare? Create a histogram of
the lead values for just the first draw and comment on the histogram's
shape -- does it confirm the earlier relationships between mean and
median?
:::

```{r Summary of Lead}
lead_summary <- flint %>%
  group_by(draw) %>%
  summarise(mean_lead = mean(lead, na.rm = TRUE), median_lead = median(lead, na.rm = TRUE))
lead_summary
ggplot(flint %>% filter(draw == "first"), aes(x = lead)) + 
  geom_histogram()
```

*The histogram shows that the means from earlier correlate based off of the way the lead distribution is*

### Question 5

```{=html}
<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
```
::: blue
Compute the *85th percentile* of lead values for each draw. (Hint: use
the `quantile` function). Comment on what you observe. Is any draw above
the EPA action threshold level?
:::

```{r Quantile calculation}
lead_percentiles <- flint %>%
  group_by(draw) %>%
  summarise(percentile_85 = quantile(lead, probs = 0.85, na.rm = TRUE))

lead_percentiles
```

*I see that the draws are between 4-16 percentile for the 85th which the first and second draw are way above th 10% threshold for action*

### Question 6

```{=html}
<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
```
::: blue
Using the data from zip code **48505**, create a plot similar to the one
in the reference document for zip code 48503.

In 1-2 sentences, comment on whether the plot confirms or contradicts
the statement below, pulled from Langkjear-Bain (2017)

"Pre-stagnation flushing" -- as it is known -- "may potentially lower"
lead levels as flushing "removes water that may have been in contact
with the lead service line for extended periods"
:::

```{r boxplot zipcode 48505}

lead_clean <- flint %>%
  filter(zip == 48505 & !is.na(lead))

ggplot(lead_clean, aes(draw,lead)) +
  geom_boxplot() + labs(title = "Lead values by Draw for Zip Code 48505", x = "Sameple draw", y = "Lead (ppm)")
```

*The boxplot for ZIP code 48505 shows that lead levels are higher in the First Draw and gradually decrease in the Second and Third Draws, supporting the idea that pre-stagnation flushing helps remove contaminated water.*

### Question 7

```{=html}
<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
```
::: blue
What is the largest lead value? What draw and zip code does it belong
to? Comment on how we should handle this value if further statistical
analysis were to be performed.
:::

```{r Max lead table}
max_lead <- flint %>%
  filter(lead == max(flint$lead, na.rm = TRUE)) %>%
  select(zip, draw, lead)

max_lead
```

*The lead value of 1051 ppb in zip code 48504, second draw is extremely high compared to typical values. Given that the EPA action threshold for lead is 15 ppb, this value is concerning*

```{=html}
<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
```
::: blue
What is the smallest lead value? What draw and zip code does it belong
to?
:::

```{r Minimum Lead Value}
min_lead <- flint %>%
  filter(lead == min(flint$lead, na.rm = TRUE)) %>%
  select(zip, draw, lead)

min_lead
```

### Question 8

```{=html}
<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
```
::: blue
One way to detect outliers in a dataset is with z-scores. The z-score of
a data value $x$ belonging to a sample with mean $\overline{x}$ and
standard deviation $s$ is $\frac{x - \overline{x}}{s}$, and it
represents the number of standard deviations the data value is from the
mean.

Group the data by draw and add a new variable to the data set
representing the z-score of the lead values in each draw. Check the
reference file to see whether the observations match.
:::

```{r Zscore calculations}

flint_zscores <- flint %>%
  group_by(draw) %>%
  mutate(z_score = (lead - mean(lead, na.rm = TRUE)) / sd(lead, na.rm = TRUE))

flint_zscores
```

How many z-scores exceed three in absolute value for each draw?

```{r Exceed values}

outlier_counts <- flint_zscores %>%
  filter(abs(z_score) > 3) %>% 
  group_by(draw) %>%
  summarise(outlier_count = n())

outlier_counts
```

### Question 9

```{=html}
<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
```
::: blue
Based on your analysis in questions 1-8, does it seem that flushing the
water decreases the lead content? You may include further code and
visualizations.
:::

*Flushing appears to reduce lead content in the water, as shown by the decreasing mean, median, and 85th percentile lead levels from the First Draw to the Third Draw. The number of extreme outliers also drops in later draws, suggesting that for most households, flushing helps lower lead exposure. However, some locations, like zip 48504, still show high lead levels even after flushing, indicating that in certain cases, flushing alone may not be enough.*
# Essential details

### Deadline and submission

-   The deadline to submit Homework 2 is **11:00pm on Friday, March
    14th, 2025.** This is a individual assignment.

-   Submit your work by uploading your RMD and HTML/PDF files through
    D2L. Kindly **double check your submission to note whether the
    everything is displayed in the uploaded version of the output in D2L
    or not.** If submitting HTML outputs, please zip the .rmd and html
    files for submission.

-   Please enable the **`echo=TRUE`** option in your code chunks and
    name each code chunk.

-   **Late work will not be accepted except under certain extraordinary
    circumstances or if an extension has been approved prior to the
    deadline.**

### Help

-   Feel free to visit us in during our virtual office hours or make an
    appointment.

-   Communicate with your classmates, but do not share snippets of code.

-   Start the assignment early. **If you wait until the last minute to
    start and try to ask questions the night the assignment is due, we
    cannot guarantee a timely response.**

### Academic integrity

This is an individual assignment. You may discuss ideas, how to debug
code, and how to approach a problem with your classmates. You may not
copy-and-paste another's code from this class or copy answers directly
from an AI tool (example: ChatGPT). As a reminder, below is the policy
on sharing and using other's code.

> Similar reproducible examples (reprex) exist online that will help you
> answer many of the questions posed on group assignments, and homework
> assignments. Use of these resources is allowed unless it is written
> explicitly on the assignment. While you may use AI tools to help debug
> code or understand a topic, you may not use it to answer assignment
> questions for you. You must ALWAYS cite any code you copy or use as
> inspiration. Copied code without citation is plagiarism and will
> result in a 0 for the assignment.

### Grading

Use the R Markdown blank file that is provided. If you want, you can use
your own formatting. Self-formatting is at your discretion but is
graded. Use the in-class assignments and resources available online for
inspiration. Another useful resource for R Markdown formatting is
available at: <https://holtzy.github.io/Pimp-my-rmd/>

| **Topic**                          | **Points** |
|------------------------------------|-----------:|
| Questions(total 15)                |         75 |
| R Markdown formatting and knitting |          7 |
| Communication of results           |         10 |
| Code style                         |          8 |
| **Total**                          |    **100** |

Please note: Code style includes code efficiency.

# Reference

1.  <http://www.cdc.gov/nchs/data_access/Vitalstatsonline.htm>

2.  Langkjr-Bain, R. (2017), The murky tale of Flint's deceptive water
    data. Significance, 14: 16-21.

3.  <https://holtzy.github.io/Pimp-my-rmd/>
