---
title: "Monte Carlo simulations"
author: "<13>"
date: "M5 ICA1"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: false
    df_print: paged
---

<style type="text/css">
/* Title */
h1.title {
  color: #262626;
  font-size:40px;
  font-weight: bold;
}
/* Level 1 header */
h1 {
  color: #71C144;
}
/* Level 2 header */
h2 {
  color: #FF664D;
}
/* Table of contents */
.list-group-item.active, .list-group-item.active:focus, .list-group-item.active:hover {
    z-index: 2;
    color: #fff;
    background-color: #71C144;
    border-color: #337ab7;
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE, 
                      warning = FALSE,
                      comment = NA)
set.seed(2738952)
```

# Introduction

>Monte Carlo simulations are a statistical technique used to model 
probabilistic (or "stochastic") processes and establish the odds for a
variety of outcomes. The concept was first popularized right after 
World War II. To study nuclear fission, mathematician Stanislaw Ulam coined 
the term in reference to an uncle who loved playing the odds at the Monte Carlo casino (then a world symbol of gambling, like Las Vegas today).

To get started, load package `tidyverse` and set the random number generator
in your setup R chunk. In the following problems you will make heavy use of
functions

- `sample(x, size, replace = FALSE, prob = NULL)`,

- `replicate(n, expr, simplify = "array")`.

```{r packages}
library(tidyverse)
```

# Antoine Gombaud's question

Conduct a simulation to answer the question below that was initially posed by 
Antoine Gombaud (a famous gambler in the 17th century).

Which is more likely:

 a. getting at least one 6 when rolling a single fair six-sided die 4 times,
 b. getting at least one pair of sixes when two fair six-sided dice are 
    thrown 24 times.
 
The number of simulation replications you choose is at your discretion, 
but if you choose a number too small the results will not be accurate.


```{r antoines}
 #Number of simulations
n_sim <- 10000


sim_A <- replicate(n_sim, {
  rolls <- sample(1:6, 4, replace = TRUE)
  any(rolls == 6)  
})

```

```{r antoine}

sim_B <- replicate(n_sim, {
  rolls <- replicate(24, sample(1:6, 2, replace = TRUE))  
  six_pairs <- sum(rowSums(rolls == 6) == 2)  
  six_pairs >= 1  
})

prob_A <- mean(sim_A)  
prob_B <- mean(sim_B) 

prob_A
prob_B
```

# Enough?

How many Monte Carlo experiments are enough? 

 a. Perform Monte Carlo simulation to evaluate the probability of getting "Heads" in a fair coin toss. 
 
 b. Use `ggplot()` to plot the probability estimate on the y-axis and the iteration (number of Monte Carlo) experiments on the x-axis. As the number of iterations gets large, you should see this value stabilize at around 0.50.

```{r enough}


flip <- function(n = 1) {
  sample(c("H", "T"), size = n, replace = TRUE)
}

set.seed(42)  
toss.df <- data.frame(toss = unlist(replicate(10000, flip(1)), use.names = FALSE))


toss.df <- toss.df %>%
  mutate(iterations = row_number(),
         heads = ifelse(toss == "H", 1, 0),  
         cum.prob = cumsum(heads) / iterations)  

ggplot(toss.df, aes(x = iterations, y = cum.prob)) +
  geom_line()
```

# Birthday problem

Conduct a Monte Carlo simulation to answer the following questions 
related to the birthday problem or birthday paradox.

  a. What is the probability of at least two people sharing the same birthday 
     (month and day) from a random sample of 23 individuals?
  b. What is the probability of at least two people sharing the same birthday 
     (month and day) from a random sample of 70 individuals?
  c. Create a plot using `ggplot` with the number of individuals on the x-axis 
     and the probability of at least one pair on the y-axis. You should 
     simulate the probability of at least two people sharing the same 
     birthday for each number of individuals from 2 to 100.
  
You may ignore leap year and assume 365 days per year. The number of simulation replications you choose is at your discretion, but if you choose a number 
too small the results will not be accurate.


```{r birthday-problem}
#create a fuction bday.match(n) which will check whether there is a birthday match in a year between `n` individuals

bday.match <- function(n) {
  birthdays <- sample(1:365, n, replace = TRUE)  # Generate n random birthdays
  any(duplicated(birthdays))  # Check if there's at least one duplicate
}
# For (a) check probability of match for 23 people
emp.23 <- mean(replicate(100000, bday.match(23)))
emp.23

#(b)
emp.70 <- mean(replicate(n_sim, bday.match(70)))
# Here're the steps to create the dataframe needed for part (c)
nmat <- matrix(c(2:100))
event.results <- replicate(10000, apply(nmat, 1, bday.match))
emp.results <- apply(event.results, 1, mean)
emp.results.df <- data.frame(n = nmat, emp.prob = emp.results)


ggplot(emp.results.df, aes(x = n, y = emp.prob)) +
  geom_line(color = "blue")


```

# References

1. http://news.mit.edu/2010/exp-monte-carlo-0517
2. https://en.wikipedia.org/wiki/Antoine_Gombaud